### 2022-07-11 bools/bools.tex
geoff ch 5

In Chapter~\ref{ch:intro} we motivated our work by the need to bridge the gap
between higher-order frontends and first-order backends. This gap is
traditionally bridged using translations from higher-order to
first-order logic \cite{ar-70-hol, mp-08-trans}. However, as shown in Chapter
\ref{ch:ehoh}, translations are usually less efficient than native support. The
distinguishing features of higher-order logic used by proof assistants that the
translation must eliminate include $\lambda$ binders, function
extensionality---the property that functions are equal if they agree on every
argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x
\, z \ieq y \,z) \iimplies x \ieq y$---and formulas occurring as arguments of
function symbols \cite{mp-08-trans}.

@@

% A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
% extension of superposition for extensional Boolean-free higher-order logic. The
% extension removes the need to translate $\lambda$-binders and functional
% extensionality. 
As mentioned in Sect.~\ref{sec:pre:ho-sup-calculi}, Bentkamp et al. developed
\lsup{}, a complete higher-order calculus that alleviates the need to translate
$\lambda$-terms and function extensionality. Kotelnikov et al.\
\cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
first-order logic to support the third feature of higher-order logic that
requires translation. They described two approaches: One based on a calculus-level
treatment of Booleans and other, which requires no changes to the calculus,
based on preprocessing.

\looseness=-1 To fully bridge the gap between higher-order and first-order tools,
we combine the two approaches: We take \lsup{} as our basis and  extend it with inference rules that reason with
Boolean terms. In early work, Kotelnikov et al.\ 
\cite{kotelnikov-15-fool} have described a \emph{FOOL paramodulation}
rule that, under some order requirements, removes the need for the axiom
describing the Boolean domain---$\iforall p. \; p \ieq \itrue \llor p
\ieq \ifalse$. In this approach, it is assumed that a problem with formulas occurring as
arguments of symbols is translated to first-order logic.

@@

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the higher-order provers
Leo-III \cite{sb-21-leo3} and Satallax \cite{cb-12-satallax}, such as
elimination of Leibniz equality, primitive instantiation, and treatment of the choice
operator \cite{pa-01-classical-ty-thy}. We have also designed new rules including those that use
higher-order unification to resolve Boolean formulas that are hoisted to the literal
level, delay clausification of nonatomic literals, and reason about formulas under
$\lambda$-binders. Even though the rules we use are inspired by
the ones of refutationally complete higher-order provers, we do not guarantee
completeness of our extension of \lsup{}. Using the insights that we gained with
this incomplete extension of \lsup{} to full higher-order logic, Bentkamp et al.
designed a complete calculus for full higher-order logic---\osup{}
\cite{bbtv-21-full-ho-sup}.

\looseness=-1
We compare our native approach with two alternatives which are based on
preprocessing (Sect.~\ref{sect:bool:alternative}). First, we compare it to  an
axiomatization of the theory of Booleans. Second, inspired by work of
Kotelnikov et al.\ \cite{kotelnikov-16-fool}, we implement the preprocessing approach that does not
require introduction of Boolean axioms.
We discuss some examples, coming from TPTP \cite{gs-17-tptp}, that
illustrate advantages and disadvantages of our approach (Sect.~\ref{sect:bool:examples}).

@@

We base our work (and parts of the following text) on Bentkamp et al.'s
\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. This
logic supports rank-1 polymorphism, as defined by the TH1 format of TPTP
\cite{ksr-16-th1}.
We
extend the syntax of this logic by adding logical connectives to the signature.
The semantics of the logic is extended by interpreting Boolean type $o$ as a
two-element domain. This amounts to extending the logic to full higher-order
logic. Taking a different perspective, this logic is an extension of the one
presented in Chapter \ref{ch:ehoh} with $\lambda$-abstraction, polymorphic
types, and native logical connectives. For reference, we provide a
description of this polymorphic logic. For the notions that are not defined here,
we assume the definitions introduced in Chapter~\ref{ch:pre}.

@@

As mentioned in the previous chapter, Zipperposition is an open source prover
written in OCaml. From its inception, it was designed as a prover that supports
easy extension of its base superposition calculus to various theories, including
arithmetic, induction, and limited support for higher-order logic
\cite{sc-15-simon-phd,sc-supind-17}.

In Zipperposition, applications are represented in flattened, spine notation (Sect.~\ref{sec:ehoh:types-and-terms}).
This means that the higher-order term $(\cst{f} \, \cst{a}) \, \cst{b}$ is
represented the same as the first-order term $\cst{f}(\cst{a}, \cst{b})$. In
addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
the nested applications of these symbols. For example, terms $\cst{p} \iand
(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. Zipperposition's
support for $\lambda$-terms is used to represent quantified nested Booleans:
Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of
the input problem, no nested Booleans will be modified or renamed using fresh
predicate symbols.

@@

Our decision to represent negative atoms as negative equations was motivated by
the need to alter Zipperposition's earlier behavior as little as possible. 
Namely, negative atoms were not used as literals that can be used
to paramodulate from, and as such added to the laziness of the superposition calculus.
However, it might be useful to consider unit clauses of the form $f \noteq \itrue$
as $f \eq \ifalse$ to strengthen rewriting. To that end, we have introduced the following
rule:
%
$$ \namedsimp{BoolDemod}{f \noteq \itrue \qquad C[\sigmaterm{f}]}{f \noteq \itrue \qquad C[\ifalse]} $$

@@

\looseness=-1
The axiom of choice is notoriously difficult to handle efficiently in higher-order provers. Andrews formulates
this axiom as $\iforall p.\, (\iexists
x.\, p \, x) \iimplies p \, (\varepsilon \, p)$, where $\varepsilon :
\forallty{\alpha}(\alpha \rightarrow o) \rightarrow \alpha$ denotes the \emph{choice
operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
\, (\varepsilon \, P) \eq \itrue$. Since the term $P \, X$ matches any Boolean
term in the proof state, this axiom is very explosive. Therefore, Leo-III
\cite{sb-21-leo3} heuristically recognizes symbols that correspond to choice and deals with
them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
added to the set of choice functions $\mathit{CF}$ (which initially contains just
$\varepsilon$). Later, elements of $\mathit{CF}$ will be used to heuristically
instantiate the axiom of choice. We reused the method of recognizing choice
functions, but generalized the rule for creating the instance of
the axiom (assuming $\xi \in \mathit{CF}$ and $X$ and $z$ are fresh variables):
%
$$\namedinference{Choice}
               {C[\xi \, t]}
               { X \, (t \, Y) \noteq \itrue \llor X \, (t \, (\xi \, (\lam{z}{X \, (t \, z)})) ) \eq \itrue }$$
%
Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
an arbitrary context around $t$, the chosen instantiation for $P$ from the axiom of choice;
the variable $X$ can later be replaced by imitations of logical symbols to create more
complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, depending on
Zipperposition options, $D$ will either be deleted or kept. Note that $D$
will not subsume its instances, as the matching algorithm Zipperposition uses favors performance over completeness and is
thus too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
\looseness=-1

@@

\looseness=-1
Consider the clause $\cst{f} \, (\inot\cst{p} \ior \inot\cst{q}) \noteq \cst{f} \,
(\inot (\cst{p} \iand \cst{q}))$. This problem is unsatisfiable, as
the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
equal. Without the \infname{Abs} rules Zipperposition relies on the
\infname{Cases}(\infname{Simp}) and \infname{IC} rules to derive the empty
clause. On the other hand, \infname{AbsER} generates~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
\inot (\cst{p} \iand \cst{q})$. Then, $C$ gets clausified using
\infname{IC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
\inot\cst{q} \iequiv \inot (\cst{p} \iand \cst{q}))$, which is first-order.

\begin{sloppypar}
  Zipperposition restricts \infname{AbsSup} by
  requiring that $s$ and $s'$ are not of function or Boolean type. If the terms are of function type, our experience is
  that a better treatment of function extensionality is to apply fresh free
  variables (or Skolem terms, depending on the sign
  \cite{bbtvw-21-sup-lam}) to both sides of a (dis)equation to reduce it to
  a first-order literal; Boolean extensionality is usually better supported by
  applying \infname{IC} on the top-level Boolean term. Thus, for the following
  discussion we can assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
  and $s'$ have the same head, and a functional or Boolean subterm. To speed up
  retrieval of such terms, we added an index that maps symbols to positions in
  clauses where they appear as a head of a term that has a functional or Boolean
  subterm. This index will be empty for first-order problems, incurring no
  overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply the \infname{Abs} rules if all
  disagreement pairs have at least one side whose head is a variable; those will
  be dealt with more efficiently using the core rules of the superposition
  calculus. To simplify the resulting clauses of \infname{Abs} rules, we also
  eagerly remove literals $s_i \not\eq s'_i$ using a simplifying version of
  \infname{ER} (Sect.~\ref{sec:pre:rules}). In particular, we apply the first
  unifier returned by the terminating, pragmatic variant of unification
  algorithm described in Sect.~\ref{sec:unif:the-unification-procedure}, and
  remove the premise after the conclusion of \infname{ER} has been computed.
\end{sloppypar}

@@

\ourpara{Preprocessing Booleans}
\looseness=-1
Kotelnikov et al.\ extended VCNF, Vampire's algorithm for clausification, to support
nested Booleans \cite{kotelnikov-16-fool}. As explained in Sect.~\ref{sec:ehoh:preprocessing}, we extended
the clausification algorithm of Ehoh to support nested Booleans inspired by this VCNF extension.
Zipperposition and Ehoh share the
same clausification algorithm, enabling us to reuse the extension, with one
notable difference:
Unlike in Ehoh, not all nested Booleans different from variables, $\itrue$ and $\ifalse$
will be removed. Namely, Booleans that are below $\lambda$-abstraction
and contain $\lambda$-bound variables will not be preprocessed. They cannot be easily hoisted to the level of an atom in which
they appear, since this process might leak any variables bound in the context
in which the nested Boolean appears. Similar preprocessing techniques are used in other higher-order provers
\cite{wskb-16-effective-norm}.

@@

During most of the 2010s, the core calculus of the best performing
higher-order prover at CASC was tableaux---a striking contrast from the first-order part
of the competition dominated by superposition-based provers. TPTP
problem \verb|SET557^1| might shed some light on why tableaux-based provers
excel on higher-order problems.
This problem conjectures that there is
no surjection from a set to its power set: 
$$ \neg (\iexists x. \, \iforall y. \,
\iexists z. \, x \, z \eq y) $$ 
After negating the conjecture and
clausification this problem becomes $ \cst{sk}_1 \, (\cst{sk}_2 \, Y) \eq Y   $
where $\cst{sk}_1$ and $\cst{sk}_2$ are Skolem symbols. Then, we can use the
\infname{ArgCong} rule \cite{bbtvw-21-sup-lam} which applies
fresh variable $W$ to both sides of the equation, yielding clause $ C = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W
\eq Y \, W$. Superposition-based higher-order theorem
provers (such as Leo-III, Vampire, and Zipperposition) will split this clause
into two clauses $C_1 = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \not\eq \top \llor Y
\, W \eq \top $ and $ C_2 =  \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \eq \top \llor
Y \, W \not\eq \top $. This clausification step makes the problem
considerably harder. Namely, the clause $C$ instantiated with the substitution $
\{ Y \mapsto \lamx{\inot (\cst{sk}_1 \, x \, x)}, W \mapsto \cst{sk}_2 \,
(\lamx{\inot (\cst{sk}_1 \, x \, x))} \} $ yields the empty clause. However,
if the original clause is split into two as described above, Zipperposition will
rely on the \infname{PI} rule to instantiate $Y$ with an imitation of $\inot$ and on
equality factoring to further instantiate this approximation. These desired
inferences need to be applied on both new clauses and represent only a fraction
of inferences that can be done with $C_1$ and $C_2$, reducing the chance of a
successful proof attempt. Rule \infname{BoolER} imitates the behavior of
a tableaux prover: It essentially rewrites the clause $C$ into $ \inot (\cst{sk}_1 \,
(\cst{sk}_2 \, Y) \, W) \not\eq Y \, W $, which makes finding the necessary substitution easy and
does not require a clausification step.

\looseness=-1
Combining the rule \infname{(Bool)ER} with dynamic clausification is very fruitful, as
the benchmark \texttt{SYO033\^{}1} illustrates. This problem contains the
single conjecture
$$ \iexists x. \, \iforall y.  \, x \, y \iequiv (\iforall z. \, y \,z)$$ The
problem is easily solved if we replace the variable $x$ with the constant $\forall$.
Moreover, the prover does not have to blindly guess this instantiation. Instead,
pretending that bound variables are free, it can obtain it by unifying $  X \, Y
$ with $ \iforall \, Y$ (which is the $\eta$-short form of $\iforall z. \, Y
\,z$). However, when the problem is clausified, all quantifiers are removed.
Then, Zipperposition finds the proof only if an appropriate instantiation mode of
\infname{PI} is used, and if both clauses resulting from clausifying the negated
conjecture are appropriately instantiated. In contrast, dynamic clausification will
derive the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, (\cst{sk} \, X) $
from the negated conjecture in three steps. Then, equality resolution results in
an empty clause, swiftly finishing the proof without any explosive inferences.
This effect is even more pronounced on problems \verb|SYO287^5| and
\verb|SYO288^5|, in which a critical proof step consists of instantiating a variable
with imitations of $\ior$ and $\iand$. In configurations that do not use dynamic
clausification and \infname{BoolER}, Zipperposition times out in any reasonable
time limit; with those two options it solves the mentioned problems in less than
100\,ms.

@@

\ourpara{Evaluation of the Rules}
\looseness=-1
For this part of the evaluation, we fixed a single well-perform\-ing Zipperposition configuration
called \emph{base} (b). Since we are testing a single configuration, we
used the CPU time limit of \NumberOK{15} s---roughly the time a single configuration is
given in a portfolio mode when participating in CASC. Configuration {b} uses the pragmatic variant
pv$_{1121}^{2}$ (Sect.~\ref{sec:unif:evaluation}) of the unification algorithm described in Chapter~\ref{ch:unif}. 
It enables the \infname{BoolSimp} rule,
the \infname{EC} rule, the \infname{PI} rule in \emph{Pragmatic} mode with $k=2$,
rules \infname{ElimLeibniz} and \infname{ElimPredVar}, the \infname{BoolER} rule, and the \infname{BoolEF} rules. 
To evaluate the usefulness of
all rules described above, we enabled, disabled, or changed the
parameters of a single rule, while keeping all other parameters of b intact. 
In figures that contain sufficiently many different configurations, cells are of the form $n (m)$ where $n$ is the total
number of proved problems by a particular configuration and $m$ is the number of
unique problems that a given configuration solved, compared to the other
configurations in the same figure. Intersections of rows and columns
denote the corresponding combination of parameters.
The result for the base configuration is written in \emph{italics}; the best result
is written in \textbf{bold}.

@@

\looseness=-1
Second, we evaluated all the modes of the \infname{PI} rule with three values for
parameter $k$: 1, 2, and~8 (Figure \ref{fig:pi}). The columns denote, from left to
right: disabling the \infname{PI} rule, \emph{Pragmatic} mode, \emph{Full} mode,
and \emph{Imit}$_\star$ modes with appropriate logical symbols. The rows denote
different values of $k$. The results show that different values for $k$ have a
modest effect on the success rate. The raw data reveal that when we focus our
attention to configurations with $k=2$, mode \emph{Full} can solve ten problems
no other mode (including disabling the \infname{PI} rule) can. Modes
\emph{Imit}$_{\iand}$ and \emph{Pragmatic} solve two  problems whereas
\emph{Imit}$_{\ior}$ solves one problem uniquely. This result suggests that, even
though this is not evident from Figure \ref{fig:pi}, the sets of problems solved 
by different modes somewhat differ.

>>>>

In Chapter~\ref{ch:intro} I motivated the work presented in this thesis by the need to bridge the gap
between higher-order frontends and first-order backends. This gap is
traditionally bridged using translations from higher-order to
first-order logic \cite{ar-70-hol, mp-08-trans}. However, as shown in Chapter
\ref{ch:ehoh}, translations are usually less efficient than native support. The
distinguishing features of higher-order logic used by proof assistants that the
translation must eliminate include $\lambda$ binders, function
extensionality---the property that functions are equal if they agree on every
argument, described by the axiom $\iforall xy. \allowbreak\, (\iforall z. \, x
\, z \ieq y \,z) \iimplies x \ieq y$---and formulas occurring as arguments of
function symbols \cite{mp-08-trans}.

@@

% A group of authors including Vukmirovi\'c\ \cite{bbtvw-21-sup-lam} designed an
% extension of superposition for extensional Boolean-free higher-order logic. The
% extension removes the need to translate $\lambda$-binders and functional
% extensionality. 
As mentioned in Sect.~\ref{sec:pre:ho-sup-calculi}, Bentkamp et al. developed
\lsup{}, a complete higher-order calculus that alleviates the need to translate
$\lambda$-terms and function extensionality. Kotelnikov et al.\
\cite{kotelnikov-15-fool,kotelnikov-16-fool} extended the language of
first-order logic to support the third feature of higher-order logic that
requires translation. They described two approaches: One based on a calculus-level
treatment of Booleans and another that requires no changes to the calculus,
based on preprocessing.

\looseness=-1 To fully bridge the gap between higher-order and first-order tools,
we combine the two approaches: We take \lsup{} as our basis and  extend it with inference rules that reason with
Boolean terms. In early work, Kotelnikov et al.\ 
\cite{kotelnikov-15-fool} have described a \emph{FOOL paramodulation}
rule that, under some order requirements, removes the need for the axiom
describing the Boolean domain---$\iforall p. \; p \ieq \itrue \llor p
\ieq \ifalse$. In this approach, it is assumed that a problem with formulas occurring as
arguments of symbols is translated to first-order logic using a translation they describe.

@@

\looseness=-1
Rules that extend the two orthogonal approaches form the basis of our support
for Boolean reasoning (Sect.~\ref{sect:bool:native}). In addition, we have
implemented rules that are inspired by the ones used in the higher-order provers
Leo-III \cite{sb-21-leo3} and Satallax \cite{cb-12-satallax}, such as
elimination of Leibniz equality, primitive instantiation, and treatment of the choice
operator \cite{pa-01-classical-ty-thy}. We have also designed new rules including those that use
higher-order unification to resolve Boolean formulas that are hoisted to the literal
level, delay clausification of nonatomic literals, and reason about formulas under
$\lambda$-binders. Even though the rules are inspired by
the ones of refutationally complete higher-order provers, we do not guarantee
completeness of our extension of \lsup{}. Using the insights that we gained with
this incomplete extension of \lsup{} to full higher-order logic, Bentkamp et al.
designed a complete calculus for full higher-order logic---\osup{}
\cite{bbtv-21-full-ho-sup}.

\looseness=-1
We compare our native approach with two alternatives that are based on
preprocessing (Sect.~\ref{sect:bool:alternative}). First, we compare it to  an
axiomatization of the theory of Booleans. Second, inspired by work of
Kotelnikov et al.\ \cite{kotelnikov-16-fool}, we implement the preprocessing approach that does not
require introduction of Boolean axioms.
We discuss some examples, coming from TPTP \cite{gs-17-tptp}, that
illustrate advantages and disadvantages of our approach (Sect.~\ref{sect:bool:examples}).

@@

We base our work (and parts of the following text) on Bentkamp et al.'s
\cite{bbtvw-21-sup-lam} extensional polymorphic clausal higher-order logic. This
logic supports rank-1 polymorphism, as defined by the TH1 format of the TPTP
\cite{ksr-16-th1}.
We
extend the syntax of this logic by adding logical connectives to the signature.
The semantics of the logic is extended by interpreting Boolean type $o$ as a
two-element domain. This amounts to extending the logic to full higher-order
logic. Taking a different perspective, this logic is an extension of the one
presented in Chapter \ref{ch:ehoh} with $\lambda$-abstraction, polymorphic
types, and native logical connectives. For reference, we provide a
description of this polymorphic logic. For the notions that are not defined here,
we assume the definitions introduced in Chapter~\ref{ch:pre}.

@@

As mentioned in Chapter \ref{ch:unif}, Zipperposition is an open source prover
written in OCaml. From its inception, it was designed as a prover that supports
easy extension of its base superposition calculus to various theories, including
arithmetic, induction, and limited support for higher-order logic
\cite{sc-15-simon-phd,sc-supind-17}.

In Zipperposition, applications are represented in flattened, spine notation (Sect.~\ref{sec:ehoh:types-and-terms}).
This means that the higher-order term $(\cst{f} \, \cst{a}) \, \cst{b}$ is
represented the same as the first-order term $\cst{f}(\cst{a}, \cst{b})$. In
addition, Zipperposition uses associativity of $\iand$ and $\ior$ to flatten out
the nested applications of these symbols. For example, terms $\cst{p} \iand
(\cst{q} \iand \cst{r})$ and $(\cst{p} \iand \cst{q}) \iand \cst{r}$ are
internally stored as $\iand \, \cst{p} \, \cst{q} \, \cst{r}$. Zipperposition's
support for $\lambda$-terms is used to represent quantified nested Booleans:
Formulas $\iforall x. \, \, f$ and $\iexists x. \, f$ are represented as
$\iforall \, (\lamx{f})$ and $\iexists \, (\lamx{f})$. After clausification of
the input problem, no nested Booleans are modified or renamed using fresh
predicate symbols.

@@

Our decision to represent negative atoms as negative equations was motivated by
the need to alter Zipperposition's earlier behavior as little as possible. 
Namely, negative atoms are not used as literals that can be used
to paramodulate from, and as such added to the laziness of the superposition calculus.
However, it might be useful to consider unit clauses of the form $f \noteq \itrue$
as $f \eq \ifalse$ to strengthen rewriting. To that end, we have introduced the following
rule:
%
$$ \namedsimp{BoolDemod}{f \noteq \itrue \qquad C[\sigmaterm{f}]}{f \noteq \itrue \qquad C[\ifalse]} $$

@@

\looseness=-1
The axiom of choice is notoriously difficult to handle efficiently in higher-order provers. Andrews formulates
this axiom as $\iforall p.\, (\iexists
x.\, p \, x) \iimplies p \, (\varepsilon \, p)$, where $\varepsilon :
\forallty{\alpha}(\alpha \rightarrow o) \rightarrow \alpha$ denotes the \emph{choice
operator} \cite{pa-01-classical-ty-thy}. After clausification, this axiom becomes $P \, X \noteq \itrue \llor P
\, (\varepsilon \, P) \eq \itrue$. Since the term $P \, X$ matches any Boolean
term in the proof state, this axiom is very explosive. Therefore, Leo-III
\cite{sb-21-leo3} heuristically recognizes symbols that correspond to choice and deals with
them on the calculus level. Namely, whenever a clause $C = P \, X \noteq \itrue \llor P \, (\cst{f} \,
P) \eq \itrue$ is chosen for processing, $C$ is removed from the proof state and $\cst{f}$ is
added to the set of choice functions $\mathit{CF}$ (which initially contains just
$\varepsilon$). Later, elements of $\mathit{CF}$ are used to heuristically
instantiate the axiom of choice. We reused the method of recognizing choice
functions, but generalized the rule for creating the instance of
the axiom (assuming $\xi \in \mathit{CF}$ and $X$ and $z$ are fresh variables):
%
$$\namedinference{Choice}
               {C[\xi \, t]}
               { X \, (t \, Y) \noteq \itrue \llor X \, (t \, (\xi \, (\lam{z}{X \, (t \, z)})) ) \eq \itrue }$$
%
Let $D$ be the conclusion of \infname{Choice}. The fresh variable $X$ in $D$ acts as
an arbitrary context around $t$, the chosen instantiation for $P$ from the axiom of choice;
the variable $X$ can later be replaced by imitations of logical symbols to create more
complex instantiations of the choice axiom. To generate useful instances early, we create two instances: $\substcl{\{X
\mapsto \lam{z}{z}\}}{D}$ and $\substcl{\{X \mapsto \lam{z}{\inot\, z}\}}{D}$. Then, depending on
Zipperposition options, $D$ will either be deleted or kept. Note that $D$
will not subsume its instances, as the matching algorithm Zipperposition uses favors performance over completeness and is
thus too weak for this \cite[Sect.~6]{bbtvw-21-sup-lam}
\looseness=-1

@@

\looseness=-1
Consider the clause $\cst{f} \, (\inot\cst{p} \ior \inot\cst{q}) \noteq \cst{f} \,
(\inot (\cst{p} \iand \cst{q}))$. This clause is unsatisfiable, as
the arguments of $\cst{f}$ on the different sides of the disequation are Boolean-extensionally
equal. Without the \infname{Abs} rules Zipperposition relies on the
\infname{Cases}(\infname{Simp}) and \infname{IC} rules to derive the empty
clause. On the other hand, \infname{AbsER} generates~$C = \inot\cst{p} \ior \inot\cst{q} \noteq
\inot (\cst{p} \iand \cst{q})$. Then, $C$ gets clausified using
\infname{IC}, effectively reducing the problem to $\inot (\inot\cst{p} \ior
\inot\cst{q} \iequiv \inot (\cst{p} \iand \cst{q}))$, which is first-order.

\begin{sloppypar}
  Zipperposition restricts \infname{AbsSup} by
  requiring that $s$ and $s'$ are not of function or Boolean type. If the terms are of function type, our experience is
  that a better treatment of function extensionality is to apply fresh free
  variables (or Skolem terms, depending on the sign
  \cite{bbtvw-21-sup-lam}) to both sides of a (dis)equation to reduce it to
  a first-order literal; Boolean extensionality is usually better supported by
  applying \infname{IC} on the top-level Boolean term. Thus, for the following
  discussion, we assume $s$ and $s'$ are not $\lambda$-abstractions or formulas. Then, $\infname{AbsSup}$ is applicable if $s$
  and $s'$ have the same head, and a functional or Boolean subterm. To speed up
  retrieval of such terms, we added an index that maps symbols to positions in
  clauses where they appear as a head of a term that has a functional or Boolean
  subterm. This index is empty for first-order problems, incurring no
  overhead if extensionality reasoning is not needed. One more restriction we implemented is that we do not apply the \infname{Abs} rules if all
  disagreement pairs have at least one side whose head is a variable; those will
  be dealt with more efficiently using the core rules of the superposition
  calculus. To simplify the resulting clauses of \infname{Abs} rules, we also
  eagerly remove literals $s_i \not\eq s'_i$ using a simplifying version of
  \infname{ER} (Sect.~\ref{sec:pre:rules}). In particular, we apply the first
  unifier returned by the terminating, pragmatic variant of unification
  algorithm described in Sect.~\ref{sec:unif:the-unification-procedure}, and
  remove the premise after the conclusion of \infname{ER} has been computed.
\end{sloppypar}

@@

\ourpara{Preprocessing Booleans}
\looseness=-1
Kotelnikov et al.\ extended VCNF, Vampire's algorithm for clausification, to support
nested Booleans \cite{kotelnikov-16-fool}. As explained in Sect.~\ref{sec:ehoh:preprocessing}, we extended
the clausification algorithm of Ehoh to support nested Booleans inspired by this VCNF extension.
Zipperposition and Ehoh share the
same clausification algorithm, enabling us to reuse the extension, with one
notable difference:
Unlike in Ehoh, not all nested Booleans different from variables, $\itrue$ and $\ifalse$
are removed. Namely, Booleans that are below $\lambda$-abstraction
and contain $\lambda$-bound variables will not be preprocessed. They cannot be easily hoisted to the level of an atom in which
they appear, since this process might leak any variables bound in the context
in which the nested Boolean appears. Similar preprocessing techniques are used in other higher-order provers
\cite{wskb-16-effective-norm}.

@@

During most of the 2010s, the core calculus of the best performing
higher-order prover at CASC was tableaux---a striking contrast from the first-order part
of the competition dominated by superposition-based provers. TPTP
problem \verb|SET557^1| might shed some light on why tableaux-based provers
excel on higher-order problems.
This problem conjectures that there is
no surjection from a set to its power set: 
$$ \neg (\iexists x. \, \iforall y. \,
\iexists z. \, x \, z \eq y) $$ 
After negating the conjecture and
clausification this problem becomes $ \cst{sk}_1 \, (\cst{sk}_2 \, Y) \eq Y   $
where $\cst{sk}_1$ and $\cst{sk}_2$ are Skolem symbols. Then, the
\infname{ArgCong} rule \cite{bbtvw-21-sup-lam} which applies
fresh variable $W$ to both sides of the equation can be used, yielding clause $ C = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W
\eq Y \, W$. Superposition-based higher-order theorem
provers (such as Leo-III, Vampire, and Zipperposition) split this clause
into two clauses $C_1 = \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \not\eq \top \llor Y
\, W \eq \top $ and $ C_2 =  \cst{sk}_1 \, (\cst{sk}_2 \, Y) \, W \eq \top \llor
Y \, W \not\eq \top $. This clausification step makes the problem
considerably harder. Namely, the clause $C$ instantiated with the substitution $
\{ Y \mapsto \lamx{\inot (\cst{sk}_1 \, x \, x)}, W \mapsto \cst{sk}_2 \,
(\lamx{\inot (\cst{sk}_1 \, x \, x))} \} $ yields the empty clause. However,
if the original clause is split into two as described above, Zipperposition will
rely on the \infname{PI} rule to instantiate $Y$ with an imitation of $\inot$ and on
equality factoring to further instantiate this approximation. These desired
inferences need to be applied on both new clauses and represent only a fraction
of inferences that can be done with $C_1$ and $C_2$, reducing the chance of a
successful proof attempt. Rule \infname{BoolER} imitates the behavior of
a tableaux prover: It essentially rewrites the clause $C$ into $ \inot (\cst{sk}_1 \,
(\cst{sk}_2 \, Y) \, W) \not\eq Y \, W $, which makes finding the necessary substitution easy and
does not require a clausification step.

\looseness=-1
Combining the rule \infname{(Bool)ER} with dynamic clausification is very fruitful, as
the benchmark \texttt{SYO033\^{}1} illustrates. This problem contains the
single conjecture
$$ \iexists x. \, \iforall y.  \, x \, y \iequiv (\iforall z. \, y \,z)$$ The
problem is easily solved if the variable $x$ is replaced with the constant $\forall$.
Moreover, the prover does not have to blindly guess this instantiation. Instead,
pretending that bound variables are free, it can obtain it by unifying $  X \, Y
$ with $ \iforall \, Y$ (which is the $\eta$-short form of $\iforall z. \, Y
\,z$). However, when the problem is clausified, all quantifiers are removed.
Then, Zipperposition finds the proof only if an appropriate instantiation mode of
\infname{PI} is used, and if both clauses resulting from clausifying the negated
conjecture are appropriately instantiated. In contrast, dynamic clausification
derives the clause $ X \, (\cst{sk} \, X) \not\eq \forall \, (\cst{sk} \, X) $
from the negated conjecture in three steps. Then, equality resolution results in
an empty clause, swiftly finishing the proof without any explosive inferences.
This effect is even more pronounced on problems \verb|SYO287^5| and
\verb|SYO288^5|, in which a critical proof step consists of instantiating a variable
with imitations of $\ior$ and $\iand$. In configurations that do not use dynamic
clausification and \infname{BoolER}, Zipperposition times out in any reasonable
time limit; with those two options it solves these two problems in less than
100\,ms.

@@

\ourpara{Evaluation of the Rules}
\looseness=-1
For this part of the evaluation, we fixed a single well-perform\-ing Zipperposition configuration
called \emph{base} (b). Since we are testing a single configuration, we
used the CPU time limit of \NumberOK{15} s---roughly the time a single configuration is
given in Zipperposition's portfolio mode when participating in CASC. Configuration {b} uses the pragmatic variant
pv$_{1121}^{2}$ (Sect.~\ref{sec:unif:evaluation}) of the unification algorithm described in Chapter~\ref{ch:unif}. 
It enables the \infname{BoolSimp} rule,
the \infname{EC} rule, the \infname{PI} rule in \emph{Pragmatic} mode with $k=2$,
rules \infname{ElimLeibniz} and \infname{ElimPredVar}, the \infname{BoolER} rule, and the \infname{BoolEF} rules. 
To evaluate the usefulness of
all rules described above, we enabled, disabled, or changed the
parameters of a single rule, while keeping all other parameters of b intact. 
In figures that contain sufficiently many different configurations, cells are of the form $n (m)$ where $n$ is the total
number of proved problems by a particular configuration and $m$ is the number of
unique problems that a given configuration solved, compared to the other
configurations in the same figure. Intersections of rows and columns
denote the corresponding combination of parameters.
The result for the base configuration is written in \emph{italics}; the best result
is written in \textbf{bold}.

@@

\looseness=-1
Second, we evaluated all the modes of the \infname{PI} rule with three values for
parameter $k$: 1, 2, and~8 (Figure \ref{fig:pi}). The columns denote, from left to
right: disabling the \infname{PI} rule, \emph{Pragmatic} mode, \emph{Full} mode,
and \emph{Imit}$_\star$ modes with appropriate logical symbols. The rows denote
different values of $k$. The results show that different values for $k$ have a
modest impact on the success rate. The raw data reveal that when we focus our
attention to configurations with $k=2$, mode \emph{Full} can solve ten problems
no other mode (including disabling the \infname{PI} rule) can. Modes
\emph{Imit}$_{\iand}$ and \emph{Pragmatic} solve two  problems whereas
\emph{Imit}$_{\ior}$ solves one problem uniquely. This result suggests that, even
though this is not evident from Figure \ref{fig:pi}, the sets of problems solved 
by different modes differ somewhat.
